---
title: "Solutions 4"
editor_options: 
  chunk_output_type: console
---

*Set up*

```{r}
#| code-summary: Packages et al.
#| message: false

library(tidyverse)
theme_set(theme_light(base_family = "Optima"))
```

## Exercise

**Colliders**

For this exercise I am going to ask you to create the following simulated dataset.

```{r}
N <- 1e3

d <- tibble(
  x = rnorm(N, 0, 1),
  y = rnorm(N, 0, 1)
)
```

The relationship between $x$ and $y$ should look something like this:

```{r}
d |> 
  ggplot(aes(x, y)) + 
  geom_point(shape = 21, fill = "grey90") +
  geom_ribbon(
    stat = "smooth", method = "lm", color = "black", 
    alpha = 1/4, linetype = "dashed", linewidth = 1/2
  ) +
  geom_smooth(method = "lm", show.legend = FALSE, color = "black", linewidth = 1/2) +
  coord_fixed(ratio = 1)
```

Now I am going to ask you to create an association between $x$ and $y$ via some form of collider bias.

$$
x \longrightarrow \underbrace{\text{z}}_{\small {\text{collider}}} \longleftarrow y
$$

You are tasked to do this in four different ways.

### Panel 1

```{r}
d <- d |> mutate(z = x <= 0 & y <= 0)
d <- d |> mutate(z = ifelse(as.logical(z), "yes", "no"))

d |> 
  filter(z == "no") |> 
  ggplot(aes(x, y)) + 
  geom_point() + 
  geom_point(
    data = d |> filter(z == "yes"),
    color = "grey90", alpha = 1/2
  ) +
  geom_ribbon(
    stat = "smooth", method = "lm", color = "black", 
    alpha = 1/4, linetype = "dashed", linewidth = 1/2
  ) +
  geom_smooth(method = "lm", show.legend = FALSE, color = "red", linewidth = 1/2) +
  coord_fixed(ratio = 1) +
  geom_abline(intercept = 0, slope = -1)

# ggsave("images/collider1.png", device = "png", dpi = "print", width = 5, height = 5)
```

### Panel 2

```{r}
d <- d |> 
  mutate(z = x + y <= 0) |> 
  mutate(z = ifelse(as.logical(z), "yes", "no"))

d |> 
  filter(z == "no") |> 
  ggplot(aes(x, y)) + 
  geom_point() + 
  geom_point(
    data = d |> filter(z == "yes"),
    color = "grey90", alpha = 1/2
  ) +
  geom_ribbon(
    stat = "smooth", method = "lm", color = "black", 
    alpha = 1/4, linetype = "dashed", linewidth = 1/2
  ) +
  geom_smooth(method = "lm", show.legend = FALSE, color = "red", linewidth = 1/2) +
  coord_fixed(ratio = 1) +
  geom_abline(intercept = 0, slope = -1)

# ggsave("images/collider2.png", device = "png", dpi = "print", width = 5, height = 5)

```

### Panel 3

```{r}
d <- d |> 
  mutate(z = (x + y >= -1.5) & (x + y <= 1.5)) |> 
  mutate(z = ifelse(as.logical(z), "yes", "no"))

d |> 
  filter(z == "yes") |> 
  ggplot(aes(x, y)) + 
  geom_point() + 
  geom_point(
    data = d |> filter(z == "no"),
    color = "grey90", alpha = 1/2
  ) +
  geom_ribbon(
    stat = "smooth", method = "lm", color = "black", 
    alpha = 1/4, linetype = "dashed", linewidth = 1/2
  ) +
  geom_smooth(method = "lm", show.legend = FALSE, color = "red", linewidth = 1/2) +
  coord_fixed(ratio = 1) +
  geom_abline(intercept = c(-1.5, 1.5), slope = -1)

## ggsave("images/collider3.png", device = "png", dpi = "print", width = 5, height = 5)

```

### Panel 4

```{r}
d <- d |> 
  mutate(p = plogis(2*x + 2*y)) |> 
  mutate(z = rbinom(N, 1, p)) |> 
  mutate(z = ifelse(as.logical(z), "yes", "no"))

d |> 
  ggplot(aes(x, y, fill = z)) + 
  geom_point(alpha = 3/4, shape = 21) + 
  geom_ribbon(
    stat = "smooth", method = "lm", color = "black", 
    alpha = 1/4, linetype = "dashed", linewidth = 1/2, 
    show.legend = FALSE
  ) +
  geom_smooth(method = "lm", show.legend = FALSE, color = "black", linewidth = 1/2) +
  coord_fixed(ratio = 1) +
  scale_fill_manual(values = c("grey90", "black")) +
  guides(fill = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  theme(legend.position = c(.9, .9), legend.background = element_blank())
```

## Exercise

**Conditional average treatment effect**

The average treatment effect across all units that have certain values of $X$---i.e., we are conditioning on $X$.

**Average treatment on the treated**

The average treatment effect among those who actually received the treatment. *The effect of taking away the treatment*.

$$
\text{ATT} = \operatorname{E} \big[ Y^1 - Y^0 \mid T =1 \big] = \operatorname{E} \big[ Y^1  \mid T =1 \big] - \underbrace{\operatorname{E} \big[Y^0 \mid T =1 \big]}_\text{unobservable}
$$

**Average treatment on the untreated**

The average treatment effect among those who did not actually receive the treatment. *The effect of adding treatment*.

$$
\text{ATT} = \operatorname{E} \big[ Y^1 - Y^0 \mid T = 0 \big] = \underbrace{\operatorname{E} \big[ Y^1  \mid T = 0 \big]}_\text{unobservable} - \operatorname{E} \big[Y^0 \mid T = 0 \big]
$$

## Exercise

Provide an example of a treatment effect that you would expect to be highly heterogeneous, and explain why you think it is likely to be heterogeneous.

-   *The effect of money on happiness. The effect of earnings on happiness will vary (aka interact) across a wide range of things (e.g., neighborhood, job, and many, many others).*

-   *The effects of political cues on environmental behavior. We would expect that this effect is heterogeneous, as complying with these cues depends on one’s political awareness and motivation to comply.*

## Exercise

Consider the data in the table below that shows the hypothetical treatment effect of cognitive behavioral therapy on depression for six participants. For the sake of this example, the six participants represent the population of interest.

| Case | Age |   Gender   | Effect |
|:----:|:---:|:----------:|:------:|
|  A   | 15  |    Man     |   7    |
|  B   | 40  |   Woman    |   3    |
|  C   | 30  |   Woman    |   7    |
|  D   | 20  | Non-binary |   8    |
|  E   | 15  |    Man     |   7    |
|  F   | 25  |   Woman    |   4    |

*What is the overall average treatment effect for the population?*

```{r}
mean(c(7, 3, 7, 8, 7, 4))
```

*What is the average treatment effect for Women?*

```{r}
mean(c(3, 7, 4))
```

*If nearly all Non-binary people get treated, and about half of all Women get treated, and we control for the differences between Women and Non-binary people, what kind of treatment effect average will we get, and what can we say about the numerical estimate we’ll get?*

NHK refers to variance-weighted ATE in the context of regression!

*If we assume that, in the absence of treatment, everyone would have had the same outcome, and also only teenagers (19 or younger) ever receive treatment, and we compare treated people to control people, what kind of treatment effect average will we get, and what can we say about the numerical estimate we’ll get?*

We *could* get a **conditional ATE**

## Exercise

*Give an example where the average treatment effect on the treated would be more useful to consider than the overall average treatment effect, and explain why.*

Job training program. Typically people who don’t need it, don’t receive it. And we would never think about giving these to people who don’t need it. So, there’s no sense in estimating the ATE—e.g., what’s the point of estimating a millionaire’s *counterfactual* after said job training program (it’s also probably negative).

If we have reason to believe that the distribution of treatment effects is heterogenous, then we might think twice before trying to estimate the ATE. For example, if job trainings help poor people but hurt rich people, why would we use the ATE and average across both groups???

## Exercise

Which of the following describes the average treatment effect of assigning treatment, whether or not treatment is actually received?

a.  Local average treatment effect

b.  Average treatment on the treated

c.  **Intent-to-treat (this one)**

d.  Variance-weighted average treatment effect

## Exercise

*Suppose you are conducting an experiment to see whether pricing cookies at \$1.99 versus \$2 affects the decision to purchase the cookies. The population of interest is all adults in the United States. You recruit people from your university to participate and randomize them to either see cookies priced as \$1.99 or \$2, then write down whether they purchased cookies. What kind of average treatment effect can you identify from this experiment?*

This is an ATE, conditional on people attending this particular university.

*Note. NHK will sometimes speak of a CATE.*

## Exercise

For each of the following identification strategies, what kind of treatment effect(s) is most likely to be identified?

*A randomized experiment using a representative sample*

ATE

*True randomization within only a certain demographic group*

CATE

*Closing back door paths connected to variation in treatment*

Variance-Weighted ATE—i.e., ATE using Linear Regression.

*Isolating the part of the variation in treatment variable that is driven by an exogenous variable*

LATE (you should have been able to figure this out from the textbook!). It stands from local average treatment effect.

*The control group is comparable to the treatment group, but treatment effects may be different across these groups*

Both ATU and ATT, which should allow us to estimate the ATE.
