---
title: "Week 1"
callout-appearance: simple
callout-icon: false
---

## Exercise {#sec-glossary}

::: callout-note
Start reading [*What Is Your Estimand? Defining the Target Quantity Connects Statistical Evidence to Theory*](https://journals.sagepub.com/doi/full/10.1177/00031224211004187)*.*

Keep track of at least 5 terms that are new to you or that perhaps you don't fully understand yet (e.g., estimand, DAG, potential outcome).

I will be asking you to build a glossary of terms related to causal inference. I will not grade this yet, but still do your best attempt at defining each of these terms.
:::

*We will be digesting this article more fully in the coming weeks, so don't worry if you don't understand everything yet.*

## Exercise

```{r}
#| label: Packages
#| code-summary: Packages
#| message: false

library(tidyverse)
library(gt)
```

## Exercise

The following data frame contains the potential outcomes for 8 individuals.

```{r}
d <- data.frame(
  T = c(0, 0, 1, 0, 0, 1, 1, 1),
  Y0 = c(5, 8, 5, 12, 4, 8, 4, 9),
  Y1 = c(5, 10, 3, 13, 2, 9, 1, 13), 
  id = LETTERS[1:8]
)

gt(d, rowname_col = "id") # gt is used for fancy printing of tables
```

The variable `T` depicts whether someone got the "treatment" or not.

::: callout-note
-   Create a new variable called `Y` that contains the observed outcomes.

-   What is the Average Treatment Effect (ATE) for this 8 person experiment?
:::

*Note. I'm only asking for a simple difference in means; this is also sometimes called the "naive estimator" (or "naive comparison"), which works just fine in simple experimental settings.*

## Exercise

::: callout-note
-   Simulate a new completely randomized experiment on these 8 people; that is, re sample $T$ at random so that equal numbers get the treatment and the control.
-   Create a new variable called `Y` that contains the observed outcomes.
-   What is the Average Treatment Effect (ATE) for this 8 person experiment?
-   Do this a couple of times (at least 3) and note the differences.
:::

::: callout-tip
Hint: You can re-sample $T$ very easily using the `sample()` function.

For example:

```{r}
sample(d$T)
```
:::

::: callout-note
How do these estimates compare to the "real" ATE?

When I say "the real ATE" I basically mean this:

```{r}
mean(d$Y1 - d$Y0)
```
:::

## Exercise

::: callout-note
Obviously, an experiment of 8 people will not give you enough "statistical power."

-   Assuming the ATE is $0.125$, how many people would you need to enroll in this experiment to have enough statistical power?
:::

::: callout-tip
Hint: There are a few different ways of giving a reasonable answer to this question. The wording of this problem is ambiguous.
:::
